---
title: "Exploring NEON Data"
format: markdown_github
editor: visual
---

## Finding Some Data

```{r Libraries}

#renv::restore()

library(tidyverse)
library(neonUtilities)
library(terra)
library(leaflet)
library(rstac)
```

The first step in any data analysis project is finding some data. For this project, I am going to check out NEON data, probably for their Yellowstone station.

I decided to start by just downloading some data through their data portal and having a look at it.

```{r January}

# Loads in what might be january's csv
jan_csv <- read_csv("./data/NEON_conc-h2o-soil-salinity/NEON.D12.YELL.DP1.00094.001.2024-01.basic.20250129T000730Z.RELEASE-2026/NEON.D12.YELL.DP1.00094.001.001.501.001.SWS_1_minute.2024-01.basic.20240205T000223Z.csv")

glimpse(jan_csv)

min(jan_csv$startDateTime)
max(jan_csv$startDateTime)
max(jan_csv$VSWCMean)
```

Well, January is not super helpful, which I suppose makes sense. All the water is frozen.

```{r July}

# Loads in 
july_csv <- read_csv("./data/NEON_conc-h2o-soil-salinity/NEON.D12.YELL.DP1.00094.001.2024-07.basic.20260123T000749Z.RELEASE-2026/NEON.D12.YELL.DP1.00094.001.004.501.001.SWS_1_minute.2024-07.basic.20240906T015521Z.csv")

glimpse(july_csv)
```

Okay, this looks like much more usable data. I think 501 is the sensor number, because there appears to be 6 sensors at the YELL location, 501-506. We can get more data on this by checking the sensor position file.

```{r Sensor Positions}

# Load sensor Positions
sensor_positions <- read_csv("./data/NEON_conc-h2o-soil-salinity/NEON.D12.YELL.DP1.00094.001.2024-07.basic.20260123T000749Z.RELEASE-2026/NEON.D12.YELL.DP1.00094.001.sensor_positions.20240906T015521Z.csv")

glimpse(sensor_positions)

# Filter for just the Water Content sensors
water_sensor_pos <- sensor_positions %>% 
  filter(str_detect(HOR.VER, "\\d{3}(?=.501)") == T)

# Quick plot
plot(water_sensor_pos$locationReferenceLongitude, water_sensor_pos$locationReferenceLatitude)

# Plot with satellite imagery
leaflet() %>% 
  addProviderTiles("Esri.WorldImagery") %>% 
  addMarkers(lng = water_sensor_pos$locationReferenceLongitude,
             lat = water_sensor_pos$locationReferenceLatitude,
             label = water_sensor_pos$HOR.VER)
```

Cool. This is useful. I have 4 sensors strung out in a line. I think I want to see what I can do with this data with some radar imagery. The trees make it a non-ideal soil moisture sensing situation, but let's see what we can do.

Next step is to convert these to a spatial object from which I can pull an extent to feed into RSTAC for radar data.

```{r ConvertSpat}

# Turns df into a spatVector
sensor_vect <- vect(water_sensor_pos, geom = c('locationReferenceLongitude', 
                                               'locationReferenceLatitude'), 
                    crs = 'epsg:4326', keepgeom = T, )

#Test
plot(sensor_vect, buffer = T, clip = F, box = F)
text(sensor_vect, labels = 'HOR.VER', pos = 1)
```

Okey dokey, that worked. Now let's pull a spatExtent.

```{r Extent}

# Pulling an extent
sensor_ext <- ext(sensor_vect)
```

And now for the STAC query. . . .

```{r STAC}

# Check the assets catalog for the name
stac_catalog_query <- stac(
  "https://planetarycomputer.microsoft.com/api/stac/v1"
) %>% 
  collections()

stac_catalog <- get_request(stac_catalog_query)

list_c(map(1:135, \(x) stac_catalog$collections[[x]]$id))

# Form the query
stac_query <- stac(
  "https://planetarycomputer.microsoft.com/api/stac/v1"
) %>% 
  stac_search(
    collections = "sentinel-1-rtc",
    bbox = sensor_ext[c(1, 3, 2, 4)],
    datetime = "2024-01-01/2024-12-31"
  ) %>% 
  post_request()

# Function to build the cURL
make_vsicurl_url <- function(base_url) {
  paste0(
    "/vsicurl", 
    "?pc_url_signing=yes",
    "&pc_collection=sentinel-1-rtc",
    "&url=",
    base_url
  )
}

# Building the cURLs
sentinel_urls <- make_vsicurl_url(assets_url(stac_query, "vv"))

# Retrieves just the portion of the data that falls in my bounding box
map(sentinel_urls, ~sf::gdal_utils(
  "warp",
  source = .x,
  destination = paste0("./data/sentinel_data/", str_extract(.x, "\\d{8}"), ".tif"),
  options = c(
    "-t_srs", crs(sensor_vect),
    "-te", sensor_ext[c(1, 3, 2, 4)]
  )
))
```

Now that we've got some data, lets see what it looks like.

```{r Rasters}

file_list <- list.files("./data/sentinel_data/", full.names = T)

rast_list <- map(file_list, rast)

plot(rast_list[[30]])
terra::points(sensor_vect)
```

Okay, this will be difficult. The question is whether there is enough other data from the NEON site to filter out the vegetation and keep just the soil moisture backscatter. Probably not, but maybe. Maybe if I used LiDAR data. . . .
