---
title: "Exploring NEON Data"
format: markdown_github
editor: visual
---

## Finding Some Data

```{r Libraries}

#renv::restore()

library(tidyverse)
library(neonUtilities)
library(terra)
library(leaflet)
library(rstac)
```

The first step in any data analysis project is finding some data. For this project, I am going to check out NEON data, probably for their Yellowstone station.

I decided to start by just downloading some data through their data portal and having a look at it.

```{r January}

# Loads in what might be january's csv
jan_csv <- read_csv("./data/NEON_conc-h2o-soil-salinity/NEON.D12.YELL.DP1.00094.001.2024-01.basic.20250129T000730Z.RELEASE-2026/NEON.D12.YELL.DP1.00094.001.001.501.001.SWS_1_minute.2024-01.basic.20240205T000223Z.csv")

glimpse(first_csv)

min(first_csv$startDateTime)
max(first_csv$startDateTime)
max(first_csv$VSWCMean)
```

Well, January is not super helpful, which I suppose makes sense. All the water is frozen.

```{r July}

# Loads in 
july_csv <- read_csv("./data/NEON_conc-h2o-soil-salinity/NEON.D12.YELL.DP1.00094.001.2024-07.basic.20260123T000749Z.RELEASE-2026/NEON.D12.YELL.DP1.00094.001.004.501.001.SWS_1_minute.2024-07.basic.20240906T015521Z.csv")

glimpse(july_csv)
```

Okay, this looks like much more usable data. I think 501 is the sensor number, because there appears to be 6 sensors at the YELL location, 501-506. We can get more data on this by checking the sensor position file.

```{r Sensor Positions}

# Load sensor Positions
sensor_positions <- read_csv("./data/NEON_conc-h2o-soil-salinity/NEON.D12.YELL.DP1.00094.001.2024-07.basic.20260123T000749Z.RELEASE-2026/NEON.D12.YELL.DP1.00094.001.sensor_positions.20240906T015521Z.csv")

glimpse(sensor_positions)

# Filter for just the Water Content sensors
water_sensor_pos <- sensor_positions %>% 
  filter(str_detect(HOR.VER, "\\d{3}(?=.501)") == T)

# Quick plot
plot(water_sensor_pos$locationReferenceLongitude, water_sensor_pos$locationReferenceLatitude)

# Plot with satellite imagery
leaflet() %>% 
  addProviderTiles("Esri.WorldImagery") %>% 
  addMarkers(lng = water_sensor_pos$locationReferenceLongitude,
             lat = water_sensor_pos$locationReferenceLatitude,
             label = water_sensor_pos$HOR.VER)
```

Cool. This is useful. I have 4 sensors strung out in a line. I think I want to see what I can do with this data with some radar imagery. The trees make it a non-ideal soil moisture sensing situation, but let's see what we can do.

Next step is to convert these to a spatial object from which I can pull an extent to feed into RSTAC for radar data.

```{r ConvertSpat}

# Turns df into a spatVector
sensor_vect <- vect(water_sensor_pos, geom = c('locationReferenceLongitude', 
                                               'locationReferenceLatitude'), 
                    crs = 'epsg:4326', keepgeom = T, )

#Test
plot(sensor_vect, buffer = T, clip = F, box = F)
text(sensor_vect, labels = 'HOR.VER', pos = 1)
```

Okey dokey, that worked. Now let's pull a spatExtent.

```{r Extent}

# Pulling an extent
sensor_ext <- ext(sensor_vect)
```

And now for the STAC query. . . .

```{r STAC}

# Form the query
```
